A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-18 paper/bare_conf.tex
Add code snippets: helmholtz and burgers

\section{Background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

@@

The complexity of FFC-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest (always characterized by affine memory accesses) is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{helmholtz-code}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other situations, for instance when solving a non-linear problem like Burgers (see Figure~\ref{burgers-code}), the number of arrays involved in the computation of the stiffness matrix can be much larger (14 in the case of Burgers using $p=1$ Lagrange tetrahedrons). Also, constants evaluated in outer loops, acting as scaling factors of arrays, may be required; trip counts can be larger (the increment is proportional to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

%

@@

\begin{algorithm}[h]
\caption{The PyOP2 compiler.}
\label{algo:PyOP2Compiler}
  \textbf{The PyOP2 Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper)
  \Return{ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
~~\\
// Optimize ast based on plan \\
ast.apply\_licm() \\
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  ast.vr\_tile(plan.uaj\_factor)
}
ast.padding() \\
ast.data\_align() \\
\Return{ast.from\_ast\_to\_c()}
\end{algorithm}

@@

\begin{algorithm}[h]
\caption{Procedure to estimate the most suitable unroll-and-jam factor and/or split size.}
\label{algo:applyCostModel}
  \textbf{Algorithm: ApplyCostModel}\\
  \KwIn{n\_inner\_arrays, n\_regs}
  \KwOut{uaj\_factor, sz\_split}

>>>>

\section{Background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
\label{code:helmholtz}
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
\label{code:burgers}
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//It follows 11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//It follows 10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$a hundred of sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\end{algorithm}

@@

The complexity of FFC-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest, always characterized by affine 1-stride memory accesses, is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The outer-product along the $j$ and $k$ dimensions computes the stiffness matrix $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

@@

\begin{algorithm}[t]
\caption{The PyOP2 compiler.}
\label{algo:PyOP2Compiler}
  \textbf{The PyOP2 Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper)
  \Return{ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
~~\\
// Optimize ast based on plan \\
ast.apply\_licm() \\
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  ast.vr\_tile(plan.uaj\_factor)
}
ast.padding() \\
ast.data\_align() \\
\Return{ast.from\_ast\_to\_c()}
\end{algorithm}

@@

\begin{algorithm}[t]
\caption{Procedure to estimate the most suitable unroll-and-jam factor and/or split size.}
\label{algo:applyCostModel}
  \textbf{Algorithm: ApplyCostModel}\\
  \KwIn{n\_inner\_arrays, n\_regs}
  \KwOut{uaj\_factor, sz\_split}
